Integration of national demographic-disturbance relationships into local caribou population viability projections can reduce uncertainty and inform monitoring decisions

### 2024-08-30 analysis/paper/paper.Rmd
Switch to beta distributed interannual variation, and other changes.

runScenarios = F #To do: set up to run full analysis from within this document.
doPriorCalibration=F #Note the prior calibration is time consuming.

@@

for(f in figSet){
  if(f =="EVPIall"){addBit=""}else{addBit=resultSet}
  currentFile = here::here(paste0("figs/",resultScenario,"/",f,addBit,".png"))
  newFile = here::here(paste0("figs/forPaper/",f,addBit,".png"))
  file.copy(currentFile, newFile,overwrite=T)
}

@@

Interannual variation in survival and recruitment is modelled using
truncated beta distributions [rtrunc function;
@novomestky_package_2016]:
$\dot{R}_t \sim \text{TruncatedBeta}(\bar{R}_t,\nu_R,l_R,h_R); \dot{S}_t \sim \text{TruncatedBeta}(\bar{S}_t,\nu_S,l_S,h_S)$.
Coefficients of variation among years $(\nu_R,\nu_S)$ and
maximum/minimum values $l_R,h_R,l_S,h_S$ for recruitment and survival
are given in Table A1. Expected recruitment ($\bar{R}_t$) and survival
($\bar{S}_t$) vary with disturbance according the beta regression models
estimated by Johnson et al. [-@johnson_science_2020]:

@@

```{r exampleTrajectories, fig.cap="Examples of simulated population trajectories from a model informed only by national disturbance-demographic relationships with (top row) and without (bottom row) delayed reproduction. Orange dots show means of 500 example trajectories. For reference, black lines and and shaded bands show means and 95\\% predictive intervals from Johnson et al's (2020) regression models without adjustment for misidentification bias and delayed reproduction.", cache=TRUE}

@@

Local demographic and covariate data provided as inputs to the Bayesian
population model include the observed state (dead or alive) of each
collared animal in each month ($\hat{I}_{i,t,m}$), the observed number
of adult females and calves in the recruitment surveys
($\hat{W}_t, \hat{J}_t$), the apparent number of adult females per
collared female in the composition survey $w$, and the measures of
anthropogenic disturbance and fire ($A_t, F_t$). Recruitment rate $R_t$
is estimated from the observed number of calves $\hat{J}_t$ and adult
female caribou $\hat{W}_t$ assuming a Poisson distribution:
\begin{equation}
\hat{J}_t \sim \text{Poisson}(\hat{W}_t,R_t).
  (\#eq:hatJEst)
\end{equation}

@@

In both models recruitment probability is a function of anthropogenic
disturbance and fire with a log link [@johnson_science_2020;
@stewart_climate-informed_2023; @dyson_existing_2022]. In the Bayesian
model we add a Gaussian distributed random year effect to account for
stochastic variation among years [@eacker_web-based_2019]:
\begin{equation}
\log(R_t)=\beta^R_0+\beta^R_a A_t+\beta^R_f F_t+\epsilon^R_t; \epsilon^R_t \sim \text{Normal}(0,\sigma^2_{R}).
  (\#eq:R)
\end{equation}

@@

For consistency with [@johnson_science_2020], we model survival
probability as an adjusted function of buffered anthropogenic
disturbance $A_t$ with a log link [@stewart_climate-informed_2023;
@dyson_existing_2022]. We add a Gaussian distributed random year effect
to account for stochastic variation among years
[@eacker_web-based_2019]: 
\begin{equation}
S_t=\text{min}(46 \tilde{S_t}-0.5)/45,1); \log(\tilde{S_t})=\beta^S_0+\beta^S_a A_t+\epsilon^S_t; \epsilon^S_t \sim \text{Normal}(0,\sigma^2_{S}).
  (\#eq:S)
\end{equation}

To assess sensitivity of results to the survival estimation method and
handle cases in which there is little or no survival data, we also
implemented an alternative parametric exponential survival model,
wherein the observed state (alive or dead, $\hat{I}_{i,t,m}$) of an
individual animal $i$ in year $t$ and month $m$ is Bernoulli distributed
[@schaub_integrated_2021]: 
\begin{equation}
\hat{I}_{i,t,m} \sim \text{Bernoulli}(S^{1/12}_t\hat{I}_{i,t,m-1}).
  (\#eq:hatIEst)
\end{equation}

@@

As suggested by Dunham et al. [-@dunham_strategic_2023] we also
calculate the expected value of perfect information (EVPI) and the
expected value of sample information (EVSI) [@runge_which_2011;
@yokota_value_2004]. We assign a value or utility of
$\nu_{\alpha,\psi}=1$ when the true state of the population $\psi$
matches the assigned status $\alpha$, and value of $\nu_{\alpha,\psi}=0$
when the status assessment is incorrect. This is "the value ascribed by
the decision-maker to the outcome, and is thus the measure of management
performance" [@runge_which_2011]. The expected value of current
information is the average utility of decisions informed by knowledge of
the disturbance scenario $\tau$ (Figure \@ref(fig:figDisturbance)) and
priors from the national model:
\begin{equation}
EV_{current} = max_{\alpha}\sum_\psi[P(\psi|\tau)\nu_{\alpha,\psi}],
  (\#eq:EVcurrent)
\end{equation}
where $P(\psi|\tau)$ is the prior probability that the
true state of the population is $\psi$, and each population is assigned
a status $\alpha$ informed by the prior. The expected value under
certainty (i.e. the expected value of a decision when the true state of
the population is known) is $EV_{certainty}=1$. The difference between
$EV_{certainty}$ and $EV_{current}$ gives EVPI, a measure of the maximum
possible benefit that could theoretically be gained if the system were
perfectly known.

EVSI measures expected improvement from a particular sampling strategy
as the difference between the expected value with and without additional
sample information: 
\begin{equation}
EVSI_{\tau} = E_x[max_{\alpha}\sum_\psi[P(\psi|x,\tau)\nu_{\alpha,\psi}]] - EV_{current},
  (\#eq:EVSI)
\end{equation}
where $x$ is the sample information from a particular
example trajectory. In this study, we treat simulated observations from
500 example trajectories for each disturbance and monitoring scenario
$\tau$ (Figure \@ref(fig:figDisturbance)) as samples from the
distribution of possible values of $x$, and assume each of these sample
results is equally probable. $P(\psi|x,\tau)$ is the posterior
probability that the true state of the population is $\psi$, and each
population is assigned a status $\alpha$ informed by the sample
information.

>>>>

runScenarios = F #To do: set up to run full analysis from within this document.

@@

if(runScenarios){
    source(here::here("analysis/scripts/0_nationalModelTrajectories.R"))
    source(here::here("analysis/scripts/0_bayesianModelExamples.R"))
    #create scenarios with sensitivity1.R and run in cloud
    #run viewSensitivityResults.R for S8 and S7
    #run EVSIPlots.R
}

for(f in figSet){
  if(f =="EVPIall"){addBit=""}else{addBit=resultSet}
  currentFile = here::here(paste0("figs/",resultScenario,"/",f,addBit,".png"))
  newFile = here::here(paste0("figs/forPaper/",f,addBit,".png"))
  file.copy(currentFile, newFile,overwrite=T)
}
```

@@

Interannual variation in survival and recruitment is modelled using
truncated beta distributions [rtrunc function;
@novomestky_package_2016]:
$\dot{R}_t \sim \text{TruncatedBeta}(\bar{R}_t,v_R,l_R,h_R); \dot{S}_t \sim \text{TruncatedBeta}(\bar{S}_t,v_S,l_S,h_S)$.
Coefficients of variation among years $(v_R,v_S)$ and
maximum/minimum values $l_R,h_R,l_S,h_S$ for recruitment and survival
are given in Table A1. Expected recruitment ($\bar{R}_t$) and survival
($\bar{S}_t$) vary with disturbance according the beta regression models
estimated by Johnson et al. [-@johnson_science_2020]:

@@

```{r exampleTrajectories, fig.cap="Examples of simulated population trajectories from a model informed only by national disturbance-demographic relationships with (top row) and without (bottom row) delayed reproduction. To highlight how variation among trajectories persists as disturbance changes we have reduced interannual variability ($v_S,v_R$) by 75\\% in these examples. Orange dots show means of 500 example trajectories. For reference, black lines and and shaded bands show means and 95\\% predictive intervals from Johnson et al's (2020) regression models without adjustment for misidentification bias and delayed reproduction.", cache=TRUE}

@@

Local demographic and covariate data provided as inputs to the Bayesian
population model include the observed state (dead or alive) of each
collared animal in each month ($\hat{I}_{i,t,m}$), the observed number
of adult females and calves in the recruitment surveys
($\hat{W}_t, \hat{J}_t$), the apparent number of adult females per
collared female in the composition survey $w$, and the measures of
anthropogenic disturbance and fire ($A_t, F_t$). Recruitment rate $R_t$
is estimated from the observed number of calves $\hat{J}_t$ and adult
female caribou $\hat{W}_t$ assuming a Binomial distribution:
\begin{equation}
\hat{J}_t \sim \text{Binomial}(\hat{W}_t,R_t).
  (\#eq:hatJEst)
\end{equation}

@@

In both models recruitment probability is a function of anthropogenic
disturbance and fire with a log link and Beta distributed interannual variation [@johnson_science_2020;
@stewart_climate-informed_2023; @dyson_existing_2022]. The coefficient of variation $\nu_R$ is constant, while the mean $\tilde{R_t}$ and standard deviation $\sigma_{R,t}$ vary with disturbance:
\begin{equation}
log(\tilde{R_t})=\beta^R_0+\beta^R_a A_t+\beta^R_f F_t
  (\#eq:Rtilde)
\end{equation}

\begin{equation}
R_t \sim \text{Beta}(\tilde{R_t},\sigma_{R,t});
\sigma_{R,t} = min(\nu_R \tilde{R_t},\sqrt{\tilde{R_t} (1-\tilde{R_t})})
(\#eq:R)
\end{equation}

@@

For consistency with Johnson et al. [-@johnson_science_2020], we model survival
probability as an adjusted function of buffered anthropogenic
disturbance $A_t$ with a log link and Beta distributed interannual variation with coefficient of variation $\nu_S$, mean $\tilde{S_t}$ and standard deviation $\sigma_{S,t}$ [@stewart_climate-informed_2023;
@dyson_existing_2022]: 
\begin{equation}
\tilde{S_t}=(46 e^{\beta^S_0+\beta^S_a A_t}-0.5)/45
  (\#eq:Stilde)
\end{equation}

\begin{equation}
S_t \sim \text{Beta}(\tilde{S_t},\sigma_{S,t});
\sigma_{S,t} = min(\nu_S \tilde{S_t},\sqrt{\tilde{S_t} (1-\tilde{S_t})})
(\#eq:S)
\end{equation}

To assess sensitivity of results to the survival estimation method and
handle cases in which there is little or no survival data, we also
implemented an alternative parametric exponential survival model,
wherein the observed state (alive or dead, $\hat{I}_{i,t,m}$) of an
individual animal $i$ in year $t$ and month $m$ is Bernoulli distributed: 
\begin{equation}
\hat{I}_{i,t,m} \sim \text{Bernoulli}(S^{1/12}_t\hat{I}_{i,t,m-1}).
  (\#eq:hatIEst)
\end{equation}

@@

As suggested by Dunham et al. [-@dunham_strategic_2023] we also
calculate the expected value of perfect information (EVPI) and the
expected value of sample information (EVSI) [@runge_which_2011;
@yokota_value_2004]. We assign a value or utility of
$\eta_{\alpha,\psi}=1$ when the true state of the population $\psi$
matches the assigned status $\alpha$, and value of $\eta_{\alpha,\psi}=0$
when the status assessment is incorrect. This is "the value ascribed by
the decision-maker to the outcome, and is thus the measure of management
performance" [@runge_which_2011]. The expected value of current
information is the average utility of decisions informed by knowledge of
the disturbance scenario $\tau$ (Figure \@ref(fig:figDisturbance)) and
priors from the national model:
\begin{equation}
EV_{current} = max_{\alpha}\sum_\psi[P(\psi|\tau)\eta_{\alpha,\psi}],
  (\#eq:EVcurrent)
\end{equation}
where $P(\psi|\tau)$ is the prior probability that the
true state of the population is $\psi$, and each population is assigned
a status $\alpha$ informed by the prior. The expected value under
certainty (i.e. the expected value of a decision when the true state of
the population is known) is $EV_{certainty}=1$. The difference between
$EV_{certainty}$ and $EV_{current}$ gives EVPI, a measure of the maximum
possible benefit that could theoretically be gained if the system were
perfectly known.

EVSI measures expected improvement from a particular sampling strategy
as the difference between the expected value with and without additional
sample information: 
\begin{equation}
EVSI_{\tau} = E_x[max_{\alpha}\sum_\psi[P(\psi|x,\tau)\eta_{\alpha,\psi}]] - EV_{current},
  (\#eq:EVSI)
\end{equation}
where $x$ is the sample information from a particular
example trajectory. In this study, we treat simulated observations from
500 example trajectories for each disturbance and monitoring scenario
$\tau$ (Figure \@ref(fig:figDisturbance)) as samples from the
distribution of possible values of $x$, and assume each of these sample
results is equally probable. $P(\psi|x,\tau)$ is the posterior
probability that the true state of the population is $\psi$, and each
population is assigned a status $\alpha$ informed by the sample
information.

### 2024-08-30 analysis/paper/supplement.Rmd
ongoing
Switch to beta distributed interannual variation, and other changes.

|$\sigma^2_{R}$||Variance in random effect of year on recruitment.|

@@

|$\sigma^2_{S}$||Variance in random effect of year on survival.|

@@

|$\nu_R=0.46$||Coefficient of variation in $\dot{R}_t$ among years.|
|$\nu_S=0.08696$||Coefficient of variation in $\dot{S}_t$ among years.|

@@

|$l_S=0$||Minimum value for $\dot{S}_t$.|
|$h_R=0$||Maximum value for $\dot{R}_t$.|
|$h_S=0$||Maximum value for $\dot{S}_t$.|

@@

|$\nu_{\alpha,\psi}=1 \text{ when } \alpha=\psi, 0 \text{ otherwise}$|binary|Value of classification decision.|

@@

The choice of priors for the random effects of year on survival and recruitment are less important because the random effects have no effect on expected population growth rate (Figs \@ref(fig:PriorAllFits) and \@ref(fig:PriorAllFitsR)). We set the hyperprior for the standard deviation of the survival random effect to vary uniformly between sSigmaMin = 0.00001 and sSigmaMax = 0.06 to exceed the variation among national model simulations by a moderate amount (Fig \@ref(fig:PriorAllFits)). We set the hyperprior for the standard deviation of the recruitment random effect to vary uniformly between rSigmaMin = 0.00001 and rSigmaMax = 0.2 to exceed the variation among national model simulations by a moderate amount (Fig \@ref(fig:PriorAllFitsR)).

@@

```{r priorCalibration, cache=TRUE,echo=FALSE, message=FALSE,results='hide'}
if(doPriorCalibration){
  source(here::here("analysis/scripts/PriorsFromPIs.R"))
  source(here::here("analysis/scripts/calibratePriors.R"))
}
```

>>>>

|$\nu_{R}$||Coefficient of variation in recruitment among years.|

@@

|$\nu_{S}$||Coefficient of variation in survival among years.|

@@

|$v_R=0.46$||Coefficient of variation in $\dot{R}_t$ among years.|
|$v_S=0.08696$||Coefficient of variation in $\dot{S}_t$ among years.|

@@

|$l_S=0.61$||Minimum value for $\dot{S}_t$.|
|$h_R=0.82$||Maximum value for $\dot{R}_t$.|
|$h_S=1$||Maximum value for $\dot{S}_t$.|

@@

|$\eta_{\alpha,\psi}=1 \text{ when } \alpha=\psi, 0 \text{ otherwise}$|binary|Value of classification decision.|

@@

The choice of priors for the random effects of year on survival and recruitment are less important because the random effects have little impact on expected population growth rate (Figs \@ref(fig:PriorAllFits) and \@ref(fig:PriorAllFitsR)). We set the prior for the coefficient of variation of survival among years $\nu_S$ to vary uniformly between sNuMin = 0.001 and sNuMax = 0.13 to match the variation among national model simulations (Fig \@ref(fig:PriorAllFits)). We set the prior for the coefficient of variation of recruitment of among years $\nu_S$ to vary uniformly between rNuMin = 0.001 and rNuMax = 0.7 to match the variation among national model simulations (Fig \@ref(fig:PriorAllFitsR)).

@@

```{r priorCalibration, cache=TRUE,echo=FALSE, message=FALSE,results='hide'}
if(doPriorCalibration){
  source(here::here("analysis/scripts/0_PriorsFromPIs.R"))
  source(here::here("analysis/scripts/0_calibratePriors.R"))
}
```
